---
title: "Practical 3 - Temporal Models"
execute: 
  freeze: auto
format: 
  html: 
    mainfont: "12"
  PrettyPDF-pdf:
    keep-tex: false
    number-sections: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| message: false
#| warning: false
#| purl: false
#| results: hide
#| eval: true
library(webexercises)
knitr::purl("day2_practical_3.qmd")
```

<font size="5"> **Aim of this practical:** </font>

1.  Fit temporal models for discrete time points using `inlabru`
2.  Forecasting for future observations
3.  Set penalized complexity (PC) priors
4.  Fit temporal models to non-Gaussian data

we are going to learn:

-   How to fit autoregressive and random walk models for time series data
-   How to set PC priors in `inlabru`
-   Forecast future observations using the posterior predictive density

{{< downloadthis day2_practical_3.R dname="practical_3.R" label="Download Practical 3 R script" icon="database-fill-down" type="success" >}}

```{r child="practicals/Ar1_ex.qmd"}
```

```{r child="practicals/Ar1_Rw1_ex.qmd"}
```

```{r child="practicals/Tokyo_rainfall_ex.qmd"}
```
