---
title: ""
format: 
  html:
    theme:
      light: flatly
      dark: darkly
  PrettyPDF-pdf:
    keep-tex: true
    number-sections: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

## AR(1) models in `inlabru`

In this exercise we will:

-   Simulate a time series with autocorrelated errors.

-   Fit an AR(1) process with `inlabru`

-   Visualize model predictions.

-   Forecasting for future observations

Start by loading useful libraries:

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(INLA)
library(ggplot2)
library(patchwork)
library(inlabru)     
```

Time series analysis is particularly valuable for modelling data with temporal dependence or autocorrelation, where observations taken at nearby time points tend to be more similar than those further apart.

A time series process is a stochastic process $\{X_t~|~t \in T\}$, which is a collection of random variables that are ordered in time where $T$ is the *index set* that determines the set of discrete and equally spaced time points at which the process is defined and observations are made.

Autoregressive processes allow us to account for the time dependence by regressing $X_t$ on past values $X_{t-1},\ldots,X_{t-p}$ with associated coefficients $\phi_k$ for each lag $k = 1,\ldots,p$. Thus an **autoregressive process of order** $p$, denoted AR($p$) , is given by:

$$
X_t = \phi_1 X_{t-1} + \ldots + \phi_p X_{t-p} + \varepsilon_t; ~~ \varepsilon_t \sim \mathcal{N}(0,\sigma^2_e)
$$

Consider now an univariate time series $y_t$ which evolves over time according to some autoregressive stochastic process. For example, a time series where the system follows an AR(1) process can be defined as:

$$
\begin{aligned}
y_t &\sim \mathcal{N}(\mu_t,\tau_y^{-1})\\
\eta_t &= g^{-1}(\mu_t) = \alpha + u_t \\
u_t &= \phi u_{t-1} + \delta_t ; ~~ \delta_t \sim \mathcal{N}(0,\tau_u^{-1}); ~~ t > 1 \\
u_1 &= \mathcal{N}(0,\kappa^{-1})\\
\kappa &= \tau_u (1-\phi^2)
\end{aligned}
$$

The response $y_t$ is assumed to be normal distributed with mean $\alpha + u_t$ and precision error $\tau_y$ ( here $g(\cdot)$ is just the identity link function that maps the linear predictor to the mean of the process). Then, the process $u_t$ follows an AR(1) process where $u_1$ is drawn from a stationary normal distribution such that $\kappa$ denotes the marginal precision for state $u_t$

The covariance matrix is then given by:

$$
\Sigma = \frac{\tau^{-1}_u}{1-\phi^2}
\begin{bmatrix}
1 & \phi & \phi^2 & \ldots & \phi^{n-1}\\
\phi & 1 & \phi & \ldots & \phi^{n-2} \\
\phi^2 & \phi & 1 & \ldots & \phi^{n-3} \\
\phi^{n-1} & \phi^{n-2} & \phi^{n-3} & \ldots & 1
\end{bmatrix}
$$

Notice that conditionally on $u_t$, the observed data $y_y$ are independent from$y_{t-1},y_{t-2}.y_{t-3},\ldots$, also the conditional distribution of $u_t$ is a markov chain such that $\pi(u_t|u_{t-1},u_{t-2},u_{t-3}) = \pi(u_t|u_{t-1})$. Thus, each time point is only conditionally dependent on the two closest time points:

$$
u_t|\mathbf{u}_{-t} \sim \mathcal{N}\left(\frac{\phi}{1-\phi^2}(u_{t-1}+u_{t+1}),\frac{\tau_u^{-1}}{1-\phi^2}\right)
$$

### Simulate example data

First, we simulate data from the model:

$$
\begin{aligned}
y_t &= \alpha + u_t + \varepsilon_t;~ \varepsilon_t \sim \mathcal{N}(0,\tau_y^{-1})\\
u_t &= \phi y_{t-1} + \delta_t; ~ \delta_t \sim \mathcal{N}(0,\tau_u^{-1})
\end{aligned}
$$

```{r}
set.seed(123)

phi = 0.8
tau_u = 10
marg.prec = tau_u * (1-phi^2) # ar1 in INLA is parametrized as marginal variance
u_t =  as.vector(arima.sim(list(order = c(1,0,0), ar = phi), 
                          n = 100,
                          sd=sqrt(1/tau_u)))
a = 1
tau_e = 5
epsilon_t = rnorm(100, sd = sqrt(1/tau_e))
y = a + u_t + epsilon_t


ts_dat <- data.frame(y =y , x= 1:100)

```

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 4
#| fig-height: 4

ggplot(ts_dat,aes(y=y,x=x))+geom_line()

```

### Fitting an AR(1) model with `inlabru`

**Model components**

First, we define the model components, notice that the latent field is defined by two components: the intercept $\alpha$ and the autoregressive random effects $u_t$:

```{r}
# Model components
cmp =  ~ -1 + alpha(1) + ut(x,model = "ar1")

```

The we can define the formula for the linear predictor and specify the observational model

```{r}
# Model formula
formula = y ~ alpha + ut
# Observational model
lik =  bru_obs(formula = y ~.,
            family = "gaussian",
            data = ts_dat)

```

Lastly, we fit the model using the `bru` function and compare the model estimates with the true mdeol parameters we simulate our data from.

```{r}
# fit the model
fit.ar1 = bru(cmp, lik)

# compare against the true values

data.frame(
  true = c(a,tau_e,marg.prec,phi),
  rbind(fit.ar1$summary.fixed[,c(1,3,5)],
        fit.ar1$summary.hyperpar[,c(1,3,5)])
        ) %>% round(2)
```

**Model predictions**

Here, we will predict the mean of our time series along with 95% credible intervals. Note that this interval are for the mean and not for new observations, we will cover forecasting new observations next.

```{r}
pred_ar1 = predict(fit.ar1, ts_dat, ~ alpha + ut)

ggplot(pred_ar1,aes(y=mean,x=x))+
  geom_line()+
    geom_ribbon(aes(x = x, y = mean, ymin = q0.025, ymax = q0.975),
                alpha = 0.5) +
  geom_point(aes(y=y,x=x))

```

### **Forecasting**

A common goal in time series modelling is forecasting into the future. Forecasting can be treated as a missing data problem where future values of the response variable are missing. Let $y_m$ be the missing response, then, by fitting a statistical model to the observed data $\mathbf{y}_{obs}$, we condition on its parameters to obtain the posterior predictive distribution:

$$
\pi(y_{m} \mid \mathbf{y}_{obs}) = \int \pi(y_{m}, \theta \mid \mathbf{y}_{obs})  d\theta = \int \pi(y_{m} \mid \mathbf{y}_{obs}, \theta) \pi(\theta \mid \mathbf{y}_{obs})  d\theta
$$

This distribution, which integrates over all parameter uncertainty, provides the complete probabilistic forecast for the missing values. `INLA` will automatically compute the predictive distributions for all missing values in the response. To do so, we can augment our data set by including the new time points at which the prediction will be made and setting the response value to `NA` for these new time points:

```{r}
ts.forecast <- rbind(ts_dat, 
  data.frame(y = rep(NA, 50), x = 101:150))

```

Next, we fit the `ar1` model to the new dataset so that the predictive distributions are computed:

```{r}

cmp =  ~ -1 + alpha(1) + ut(x,model = "ar1")

pred_lik =  bru_obs(formula = y ~.,
            family = "gaussian",
            data = ts.forecast)

fit.forecast = bru(cmp, pred_lik)


```

Lastly, we can draw samples from the posterior predictive distribution using the `predict` function and visualize our forecast as follows:

```{r}
pred_forecast = predict(fit.forecast, ts.forecast, ~ alpha + ut)

p1= ggplot(pred_forecast,aes(y=mean,x=x))+
  geom_line()+
    geom_ribbon(aes(x = x, y = mean, ymin = q0.025, ymax = q0.975),
                alpha = 0.5) +
  geom_point(data=ts_dat, aes(y=y,x=x))
```
