---
title: "Practical - Geostatistics"
execute: 
  warning: false
  message: false
format: 
  html:
    theme:
      light: flatly
      dark: darkly
  PrettyPDF-pdf:
    keep-tex: true
    number-sections: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| message: false
#| warning: false
#| purl: false

# load webexercises library for tasks and questions (just for a preview - the practical compiler should take care of this when compiling multiple excercises)
library(webexercises)

```

# Geostatistics

In this practical we are going to fit a geostatistical model. We will:

-   Learn how to fit a geostatistical model in `inlabru`

-   Learn how to build a mesh for the SPDE representation

-   Learn how to add spatial covariates to the model

-   Learn how to do predictions

-   Learn how to simulate from the fitted model

------------------------------------------------------------------------

Libraries to load:

```{r}
#| warning: false
#| message: false


library(dplyr)
library(INLA)
library(inlabru) 
library(sf)
library(terra)


# load some libraries to generate nice map plots
library(scico)
library(ggplot2)
library(patchwork)
library(mapview)
library(tidyterra)

```

## The data

We will explore data on the Pacific Cod (*Gadus macrocephalus*) from a trawl survey in Queen Charlotte Sound. The `pcod` dataset is available from the `sdmTMB` package and contains the presence/absence records of the Pacific Cod during each surveys along with the biomass density of Pacific cod in the area swept (kg/Km$^2$). The `qcs_grid` data contain the depth values stored as $2\times 2$ km grid for Queen Charlotte Sound.

The dataset contains presence/absence data from 2003 to 2017. In this practical we only consider year 2003.

We first load the dataset and select the year of interest

```{r}
#| message: false
#| warning: false
library(sdmTMB)

pcod_df = sdmTMB::pcod %>% filter(year==2003)
qcs_grid = sdmTMB::qcs_grid

```

Then, we create an `sf` object and assign the rough coordinate reference to it:

```{r}
#| message: false
#| warning: false
#| 
pcod_sf =   st_as_sf(pcod_df, coords = c("lon","lat"), crs = 4326)
pcod_sf = st_transform(pcod_sf,
crs = "+proj=utm +zone=9 +datum=WGS84 +no_defs +type=crs +units=km" )
```

We convert the covariate into a raster and assign the same coordinate reference:

```{r}

depth_r <- rast(qcs_grid, type = "xyz")
crs(depth_r) <- crs(pcod_sf)

```

Finally, we can plot our dataset. Note that to plot the raster we need to also upload the `tidyterra` library.

```{r}

ggplot()+ 
  geom_spatraster(data=depth_r$depth)+
  geom_sf(data=pcod_sf,aes(color=factor(present))) +
    scale_color_manual(name="Occupancy status for the Pacific Cod",
                     values = c("black","orange"),
                     labels= c("Absence","Presence"))+
  scale_fill_scico(name = "Depth",
                   palette = "nuuk",
                   na.value = "transparent" ) + xlab("") + ylab("")

```

## The model

We first fit a simple model where we consider the observation as Bernoulli and where the linear predictor contains only one intercept and the GR field defined through the SPDE approach. The model is defined as:

**Stage 1** Model for the response

$$
y(s)|\eta(s)\sim\text{Binom}(1, p(s))
$$ **Stage 2** Latent field model

$$
\eta(s) = \text{logit}(p(s)) = \beta_0 + \omega(s)
$$

with

$$
\omega(s)\sim \text{  GF with range } \rho\  \text{ and maginal variance }\ \sigma^2
$$

**Stage 3** Hyperparameters

The hyperparameters of the model are $\rho$ and $\sigma$

**NOTE** In this case the linear predictor $\eta$ consists of two components!!

### The workflow

When fitting a geostatistical model we need to go through the following steps:

1.  Build the mesh

2.  Define the SPDE representation of the spatial GF. This includes defining the priors for the range and sd of the spatial GF

3.  Define the *components* of the linear predictor. This includes the spatial GF and all potential covariates

4.  Define the observation model using the `bru_obs()` function

5.  Run the model using the `bru()` function

### 1. Building the mesh

The first task, when dealing with geostatistical models in `inlabru` is to build the mesh that covers the area of interest. For this purpose we use the function `fm_messh_2d`.

One way to build the mesh is to start from the locations where we have observations, these are contained in the dataset `pcod_sf`.

```{r}

mesh = fm_mesh_2d(loc = pcod_sf,           # Build the mesh
                  max.edge = c(10,20),     # The largest allowed triangle edge length.
                  offset = c(5,50))       # The automatic extension distance
ggplot() + gg(mesh) + geom_sf(data= pcod_sf, aes(color = factor(present)), size = 0.1) + xlab("") + ylab("")


```

As you can see from the plot above, some of the locations are very close to each other, this causes some very small triangles. This can be avoided using the option `cutoff =` which collapses the locations that are closer than a cutoff (those points are collapsed in the mesh construction but, of course, not when it come to estimaation.)

```{r}

mesh = fm_mesh_2d(loc = pcod_sf,           # Build the mesh
                  cutoff = 2,
                  max.edge = c(10,20),     # The largest allowed triangle edge length.
                  offset = c(5,50))       # The automatic extension distance
ggplot() + gg(mesh) + geom_sf(data= pcod_sf, aes(color = factor(present)), size = 0.1) + xlab("") + ylab("")

```


Key parameters in mesh construction include: `max.edge` for maximum triangle edge lengths, `offset` for inner and outer extensions (to prevent edge effects), and cutoff to avoid overly small triangles in clustered areas.

::: callout-note
**General guidelines for creating the mesh**

1.  Create triangulation meshes with `fm_mesh_2d()`
2.  Move undesired boundary effects away from the domain of interest by extending to a smooth external boundary
3.  Use a coarser resolution in the extension to reduce computational cost (`max.edge=c(inner, outer)`)
4.  Use a fine resolution (subject to available computational resources) for the domain of interest (inner correlation range) and filter out small input point clusters (0 \< `cutoff` \< inner)
5.  Coastlines and similar can be added to the domain specification in `fm_mesh_2d()` through the `boundary` argument.
:::


::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Look at the documentation for the `fm_mesh_2d` function typing

```{r}
#| eval: false

?fm_mesh_2d
```

play around with the different options and create different meshes.

The *rule of thumb* is that your mesh should be:

-   fine enough to well represent the spatial variability of your process, but not too fine in order to avoid computation burden
-   the triangles should be regular, avoid long and thin triangles.
-   The mesh should contain a buffer around your area of interest (this is what is defined in the `offset` option) in order to avoid boundary artefact in the estimated variance.
:::

### 2. Define the SPDE representation of the spatial GF

To define the SPDE representation of the spatial GF we use the function `inla.spde2.pcmatern`. This takes as input the mesh we have defined and the PC-priors definition for $\rho$ and $\sigma$ (the range and the marginal standard deviation of the field).

PC priors Gaussian Random field are defined in (Fuglstad et al. 2018). From a practical perspective for the range $\rho$ you need to define two paramters $\rho_0$ and $p_{\rho}$ such that you believe it is reasonable that

$$
P(\rho<\rho_0)=p_{\rho}
$$

while for the margianal variance $\sigma$ you need to define two parameters $\sigma_0$ and $p_{\sigma}$ such that you believe it is reasonable that

$$
P(\sigma<\sigma_0)=p_{\sigma}
$$

You can use the following function to compute and plot the prior distributions for the range and sd of the Matern field.

```{r}
dens_prior_range = function(rho_0, p_alpha)
{
  # compute the density of the PC prior for the
  # range rho of the Matern field
  # rho_0 and p_alpha are defined such that
  # P(rho<rho_0) = p_alpha
  rho = seq(0, rho_0*10, length.out =100)
  alpha1_tilde = -log(p_alpha) * rho_0
  dens_rho =  alpha1_tilde / rho^2 * exp(-alpha1_tilde / rho)
  return(data.frame(x = rho, y = dens_rho))
}

dens_prior_sd = function(sigma_0, p_sigma)
{
  # compute the density of the PC prior for the
  # sd sigma of the Matern field
  # sigma_0 and p_sigma are defined such that
  # P(sigma>sigma_0) = p_sigma
  sigma = seq(0, sigma_0*10, length.out =100)
  alpha2_tilde = -log(p_sigma)/sigma_0
  dens_sigma = alpha2_tilde* exp(-alpha2_tilde * sigma) 
  return(data.frame(x = sigma, y = dens_sigma))
}
```

Here are some alternatives for defining priors for our model

```{r}
spde_model1 =  inla.spde2.pcmatern(mesh,
                                  prior.sigma = c(.1, 0.5),
                                  prior.range = c(30, 0.5))
spde_model2 =  inla.spde2.pcmatern(mesh,
                                  prior.sigma = c(10, 0.5),
                                  prior.range = c(1000, 0.5))
spde_model3 =  inla.spde2.pcmatern(mesh,
                                  prior.sigma = c(1, 0.5),
                                  prior.range = c(100, 0.5))
```

And here we plot the different priors for the range:

```{r}
ggplot() + 
  geom_line(data = dens_prior_range(30,.5), aes(x,y, color = "model1")) +
  geom_line(data = dens_prior_range(1000,.5), aes(x,y, color = "model2")) +
  geom_line(data = dens_prior_range(100,.5), aes(x,y, color = "model3")) 
```

and for the sd:

```{r}
ggplot() + 
  geom_line(data = dens_prior_sd(1,.5), aes(x,y, color = "model1")) +
  geom_line(data = dens_prior_sd(10,.5), aes(x,y, color = "model2")) +
  geom_line(data = dens_prior_sd(.1,.5), aes(x,y, color = "model3")) 
```

::: {.callout-tip icon="false"}
## {{< bi question-octagon color=#6dc83c >}} Question

Consider the `pcod_sf`, the spatial extension and type of the data... are some of the previous choices more reasonable than others? `r mcq(c("spde_model1","spde_model2", answer = "spde_model3"))`

**NOTE** Remember that a prior should be reasonable, but the model should not totally depend on it.
:::

### 3. Define the components of the linear predictor

We have now defined a mesh and a SPDE representation of the spatial GF. We now need to define the model components:

```{r}

cmp = ~ Intercept(1) + space(geometry, model = spde_model3)
```

**NOTE** since the dataframe we use (`pcod_sf`) is an `sf` object the input in the `space()` component is the geometry of the dataset.

### 4. Define the observation model

Our data are Bernoulli distributed so we can define the observation model as:

```{r}

formula = present ~ Intercept  + space

lik = bru_obs(formula = formula, 
              data = pcod_sf, 
              family = "binomial")

```

### 5. Run the model

Finally we are ready to run the model

```{r}
fit1 = bru(cmp,lik)
```

## Extract results

### Hyperparameters

::: {.callout-warning icon="false"}
### {{< bi pencil-square color=#c8793c >}} Task

Plot the posterior densities for the range $\rho$ and the standard deviation $\sigma$ alog with the prior for both parameters.

`r hide("Take hint")`

Posterior marginals for the hyperparameters can be extracted from the fitted model as:

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
#| eval: false

fit1$marginals.hyperpar$'name of the parameter'

```

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
#| eval: false

# Extract marginal for the range

ggplot() + 
  geom_line(data = fit1$marginals.hyperpar$`Range for space`,
            aes(x,y, color = "Posterior")) +
  geom_line(data = dens_prior_range(100,.5),
            aes(x,y, color = "Prior"))


ggplot() + 
  geom_line(data = fit1$marginals.hyperpar$`Stdev for space`,
            aes(x,y, color = "Posterior")) +
  geom_line(data = dens_prior_sd(1,.5), aes(x,y, color = "Prior"))


```
:::

## Spatial prediction

We now want to extract the estimated posterior mean and sd of spatial GF. To do this we first need to define a grid of points where we want to predict. We do this using the function `fm_pixel()`, which creates a regular grid of points covering the mesh

```{r}
pxl = fm_pixels(mesh)
```

then compute the prediction for both the spatial GF and the linear predictor (spatial GF + intercept)

```{r}
preds = predict(fit1, pxl, ~data.frame(spatial = space,
                                      total = Intercept + space))
```

Finally, we can plot the maps

```{r}
ggplot() + geom_sf(data = preds$spatial,aes(color = mean)) + scale_color_scico() + ggtitle("Posterior mean")

ggplot() + geom_sf(data = preds$spatial,aes(color = sd)) + scale_color_scico() + ggtitle("Posterior sd")
```

**Note** The posterior sd is lowest at the observation points. Note how the posterior sd is inflated around the border, this is the "border effect" due to the SPDE representation.

Instead of predicting over a grid covering the whole mesh, we can limit our predictions to the points where the covariate is defined. We can do this by defining a `sf` object using coordinates in the object `depth_r`.

```{r}
pxl1 = data.frame(crds(depth_r), 
                  as.data.frame(depth_r$depth)) %>% 
       filter(!is.na(depth)) %>%
st_as_sf(coords = c("x","y"))
```

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Produce prediction over `pxl1` using the same techniques as before. Plot your results.

`r hide("Take hint")`

Add hint details here...

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false


pred_pxl1 = predict(fit1, pxl1, ~data.frame(spatial = space,
                                      total = Intercept + space))

ggplot() + geom_sf(data = pred_pxl1$total,aes(color = mean)) + scale_color_scico() + ggtitle("Posterior mean")

ggplot() + geom_sf(data = pred_pxl1$total,aes(color = sd)) + scale_color_scico() + ggtitle("Posterior sd")

```
:::

Instead of computing the posterior mean and standard deviations of the estimated surface, one can also *simulate* possible realizations of such a surface. This will give the user a better idea of the type of realized surfaces one can expect. We can do this using the function `generate()`.

```{r}
# we simulate 4 samples from the 
gens = generate(fit1, pxl1, ~ (Intercept + space),
                n.samples = 4)

pp = cbind(pxl1, gens)

pp %>% select(-depth) %>%
  pivot_longer(-geometry) %>%
    ggplot() + 
      geom_sf(aes(color = value)) +
      facet_wrap(.~name) +
        scale_color_scico(direction = -1) +
        ggtitle("Sample from the fitted model")
```

## An alternative model

We now want to check if the `depth` covatiate has an influence on the probability of presence. We do this in two different models

1.  **Model 1** The depth enters the model in a linear way. The linear predictor is then defined as:

$$
  \eta(s) = \text{logit}(p(s)) = \beta_0 + \omega(s) + \beta_1\ \text{depth}(s)
$$

2.  **Model 1** The depth enters the model in a non linear way. The linear predictor is then defined as:

$$
  \eta(s) = \text{logit}(p(s)) = \beta_0 + \omega(s) +  f(\text{depth}(s))
$$ where $f(.)$ is a smooth function. We will use a RW2 model for this.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Fit model 1. Define components, observation model and use the `bru()` function to estimate the parameters.

**Note** Use the scaled version of the covariate stored in `depth_r$depth_scaled`.

What is the liner effect of depth on the logit probability?

`r hide("Take hint")`

Add hint details here...

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false


cmp = ~ Intercept(1) + space(geometry, model = spde_model3) +
        covariate(depth_r$depth_scaled, model = "linear")

formula = present ~ Intercept  + space + covariate

lik = bru_obs(formula = formula, 
              data = pcod_sf, 
              family = "binomial")


fit2 = bru(cmp, lik)

```
:::

We now want to fit **Model 2** where we allow the effect of depth to be non-linear. To use the RW2 model we need to *group* the values of depth into distinct classe. To do this we use the function `inla.group()` which, by default, creates 20 groups. The we can fit the model as usual

```{r}

# create the grouped variable
depth_r$depth_group = inla.group(values(depth_r$depth_scaled))

# run the model
cmp = ~ Intercept(1) + space(geometry, model = spde_model3) +
        covariate(depth_r$depth_group, model = "rw2")

formula = present ~ Intercept  + space + covariate

lik = bru_obs(formula = formula, 
              data = pcod_sf, 
              family = "binomial")


fit3 = bru(cmp, lik)

# plot the estimated effect of depth

fit3$summary.random$covariate %>% 
  ggplot() + geom_line(aes(ID,mean)) + 
                                  geom_ribbon(aes(ID, ymin = `0.025quant`, 
                                                      ymax = `0.975quant`), alpha = 0.5)
```

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Create a map of predicted *probability* from Model 3. You can use a inverse logit function defined as

```{r}
#| eval: false

inv_logit = function(x) (1+exp(-x))^(-1)
```

`r hide("Take hint")`

The `predict()` function can take as input also functions of elements of the components you want to consider `r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

inv_logit = function(x) (1+exp(-x))^(-1)

pred3  = predict(fit3, pxl1, ~inv_logit(Intercept + space + covariate) )

pred3 %>% ggplot() + 
      geom_sf(aes(color = mean)) +
        scale_color_scico(direction = -1) +
        ggtitle("Sample from the fitted model")
```
:::
