---
title: "Practical 2 - GLMM"
format: 
  html:
    theme:
      light: flatly
      dark: darkly
  PrettyPDF-pdf:
    keep-tex: true
    number-sections: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(webexercises)

```

In this practical we are going to fit a Generalized (Mixed) Linear Model in `inlabru`.

We are going to:

- Fit a Poisson regression
- Change the prior distributions for the model hyperparameters (both for fixed and random effects)
- Compute and visualize posterior densities and summaries for marginal effects
 
 
 We start by loading some useful libraries 
 
```{r}
#| warning: false
#| message: false

library(dplyr)
library(INLA)
library(ggplot2)
library(patchwork)
library(inlabru)     
library(lme4)


```

In this practical we are going to analyse the dataset `grouseticks`. The data, contained  in the library `lme4`, Number of ticks on the heads of red grouse chicks sampled in the field. See `?grouseticks` for details. 

The data are analysed in the paper 

Elston et al. “Analysis of aggregation, a worked example: numbers of ticks on red grouse chicks.” Parasitology

and in this practical we will follow their analysis.



```{r}
data(grouseticks)
?grouseticks
head(grouseticks)
```

## Fitting Poisson regression

We assume that the number of ticks $y_{ijk}$ counted on chick $i$ of brood $j$ in
year $k$ follows a Poisson distribution with mean $\lambda_{ijk}$
$$
y_{ijk}|\lambda_{ijk}\sim\text{Poisson}(\lambda_{ijk})
$$
We then model log mean counts $\eta_{ijk} = \log(\lambda_{ijk})$ as a linear function of year, altitude, brood, and
individual chick within brood. 
We assume a fixed effect $\alpha_k$ of year $k$, a linear effect of altitude $x_{ij}$, 
two random effects $e_{jk}$ and $\epsilon_{ijk}$ brood and
individual within brood respectively. Thus:
$$
\eta_{ijk} = \log(\lambda_{ijk}) = \alpha_k + \beta x_{ij} + e_{jk} + \epsilon_{ijk}
$${#eq-linpred}

where $e_{jk}\sim\mathcal{N}(0,\tau^{-1}_e)$ and $\epsilon_{jk}\sim\mathcal{N}(0,\tau^{-1}_\epsilon)$.

We first fit the model using the default priors for fixed and random effects.

To fit the model we first have to create two more variable, one that indexes the combination $jk$ of brood and year and another that indexes the combination $ijk$ of individuals per brood per year.

```{r}

grouseticks  = grouseticks %>%
  group_by(BROOD, YEAR) %>%
  mutate(ij = cur_group_id()) %>%
  ungroup() %>%
  mutate(ijk = seq_along(INDEX))
```

now we can fit the model.


::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Complete the code to fit the model in @eq-linpred. You need to define the components, the likelihood and finaly use the `bru()` function to get the results

```{r}
#| echo: true
#| eval: false

cmp= ~ -1 + year(...) +
  height(...) + 
  brood_year(...) +
  brood_year_chicken(...)

lik = bru_obs(forumula = ...,
              data = ...,
              family = ...)

fit = bru(cmp, lik)

```




```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
cmp= ~ -1 + year(YEAR, model = "iid", initial = log(0.01), fixed = T) +
  height(cHEIGHT, model =  "linear") + 
  brood_year(ij , model = "iid") +
  brood_year_chicken(ijk, model= "iid")

lik = bru_obs(TICKS ~ .,
              data = grouseticks,
              family = "Poisson")

  
fit = bru(cmp, lik)
```
:::


Now we can check the results. 


::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Plot the posterior distribution of the fixed effects: $\beta$ and $\alpha_k$. 


```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

fit$marginals.fixed$height %>% ggplot() + geom_line(aes(x,y)) + ggtitle("Linear effect of altitude")

ggplot() + geom_line(data = fit$marginals.random$year[[1]], aes(x,y, color = "96") )+
  geom_line(data = fit$marginals.random$year[[2]], aes(x,y, color = "97")) +
  geom_line(data = fit$marginals.random$year[[3]], aes(x,y, color = "98")) 
  

```
:::

We now want to look at the hyperparameters of the model


::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Plot the posterior estimated of the standard deviation $\sigma_e = \sqrt{1/\tau_e}$ and $\sigma_\epsilon = \sqrt{1/\tau_\epsilon}$ of the two random effects.

`r hide("Take hint")`

You can use the function `inla.tmarginal()` or the `generate()` function

`r unhide()`

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false


# Method 1 : use the inla.tmarginal() function

sd_e = inla.tmarginal(fun = function(x)sqrt(1/x), fit$marginals.hyperpar$`Precision for brood_year`)

sd_eps = inla.tmarginal(fun = function(x)sqrt(1/x), fit$marginals.hyperpar$`Precision for brood_year_chicken`)

ggplot() + geom_line(data = sd_e, aes(x,y, color = "sd_e")) + geom_line(data = sd_eps, aes(x,y, color = "sd_epsilon"))


# Method 1 : use the generate() function


# we use this to get the correct name of the parameters
bru_names(fit)

out = generate(fit, formula = ~ data.frame(sd_e = sqrt(1/Precision_for_brood_year),
                                     sd_eps = sqrt(1/Precision_for_brood_year_chicken)),
               n.samples = 10000)

sd_sample = data.frame(sd_e = unlist(sapply(out, function(x) x[1])),
                       sd_eps = unlist(sapply(out, function(x) x[2])))
                       

ggplot() + geom_line(data = sd_e, aes(x,y, color = "sd_e")) + 
  geom_density(data = sd_sample, aes(x = sd_e, color = "sd_e"), linetype = "dashed") +
  geom_line(data = sd_eps, aes(x,y, color = "sd_epsilon"))+ 
  geom_density(data = sd_sample, aes(x = sd_eps, color = "sd_epsilon"), linetype = "dashed")


# Note that, if one is interested in hyperparameters, sampling is not always the best choice!

```
:::

# Change prior distributions

Before trying to change the priors for the hyperparameters we can check which priors are actually used in the model

```{r}
inla.priors.used(fit)
```

 
From the output we see that the precision for the linear effect of altitude is 0.001 (which means the sd is $1/\sqrt{0.001} = 31.62$)

The precisions for the random effects have a Gamma prior with parameters 1 and 5e-05.

## Change the precision for the linear effects

The precision for linear effects is set in the component definition. For example, if we want to increase the precision to 0.1 for we define the relative components as:


```{r }
#| eval: false 
#| purl: false

cmp= ~ -1 + year(YEAR, ... ) +
  height(cHEIGHT, model =  "linear", prec.linear = 0.1) + ...

```


::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Run the model again using 0.1 as default precision for the altitude slope parameter and for the `year` fixed effect

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: false
#| fig-width: 6 
#| fig-height: 4
#| fig-align: center 
#| purl: false
#| eval: true  
 
cmp2 =  ~ -1 + year(YEAR, model = "iid", initial = log(0.1), fixed = T) +
  height(cHEIGHT, model =  "linear", prec.linear = 0.1) + 
  brood_year(ij , model = "iid") +
  brood_year_chicken(ijk, model= "iid")

fit2 = bru(cmp2, lik) 
```

Note that we can use the same observation model as before since both the formula and the dataset are unchanged.
:::


**Changing the  precision  for the random effects** 

Priors on the hyperparameters of the random effects  model must be passed by defining argument `hyper` within component of interest

```{r}

# First we define the logGamma (0.01,0.01) prior 

prec.prior <- list(prec = list(prior = "loggamma",  # prior name
                               param = c(0.01, 0.01))) # prior parameters
                        

cmp3 =  ~ -1 + year(YEAR, model = "iid", initial = log(0.1), fixed = T) +
  height(cHEIGHT, model =  "linear", prec.linear = 0.1) + 
  brood_year(ij , model = "iid", hyper = prec.prior) +
  brood_year_chicken(ijk, model= "iid", hyper = prec.prior)


fit3 = bru(cmp3, lik) 

```


Note that for `model = "linear"` the value for the precision is expressed in natural scale while for `model = "iid'` (and all other random effects) the value is expressed in log scale.


### Visualizing the posterior marginals

Posterior marginal distributions of the ﬁxed effects parameters and the hyperparameters can be visualized also using the `plot()` function by calling the name of the component. For example, if want to visualize the posterior density of the slope $\beta$ we can type:

```{r}
#| fig-width: 4
#| fig-align: center
#| fig-height: 4
#| code-fold: show

plot(fit, "height")
```

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Plot the posterior marginals for  the precision of brood and individual random effect  $\pi(\tau_e|y)$

`r hide("Take hint")`

See the `summary()` output to check the names for the different model components.

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
plot(fit, "Precision for brood_year")
```
:::
