---
title: "Practical - Point Process"
execute: 
  warning: false
  message: false
format: 
  html:
    theme:
      light: flatly
      dark: darkly
  PrettyPDF-pdf:
    keep-tex: true
    number-sections: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| message: false
#| warning: false

# load webexercises library for tasks and questions (just for a preview - the practical compiler should take care of this when compiling multiple excercises)
library(webexercises)

```

# Point process models {#sec-linmodel}

In this practical we will:

-   Fit a [homogeneous Point Process](#HPP)
-   Fit a [non-homogeneous Point Process](#NHPP)
-   Fit a [LGCP (log-Gaussian Cox Process)](#LGCP)

Libraries to load:

```{r}
#| warning: false
#| message: false


library(dplyr)
library(INLA)
library(ggplot2)
library(patchwork)
library(inlabru)     
library(spatstat)
library(sf)
library(scico)
library(spatstat)
library(lubridate)
library(terra)
library(tidyterra)
```

------------------------------------------------------------------------

## The data

In this practical we consider the data `clmfires` in the `spatstat`
library.

This dataset is a record of forest fires in the Castilla-La Mancha
region of Spain between 1998 and 2007. This region covers approximately
400 by 400 kilometres. The coordinates are recorded in kilometres. For
more info about the data you can type:

```{r}
#| echo: true
#| eval: false
#| purl: false
?clmfires
```

We first read the data and transform them into an `sf` object. We also
create a polygon that represents the border of the Castilla-La Mancha
region. We select the data for year 2004 and only those fires caused by
lightning.

```{r}
#| label: fig-points
#| fig-cap: "Distribution of the observed forest fires caused by lightning in Castilla-La Mancha in 2004"
#| 
data("clmfires")
pp = st_as_sf(as.data.frame(clmfires) %>%
                mutate(x = x, 
                       y = y),
              coords = c("x","y"),
              crs = NA) %>%
  filter(cause == "lightning",
         year(date) == 2004)

poly = as.data.frame(clmfires$window$bdry[[1]]) %>%
  mutate(ID = 1)

region = poly %>% 
  st_as_sf(coords = c("x", "y"), crs = NA) %>% 
  dplyr::group_by(ID) %>% 
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON") 
  
ggplot() + geom_sf(data = region, alpha = 0) + geom_sf(data = pp)  
```

## Fit a homogeneous Poisson Process {#HPP}

As a first exercise we are going to fit a homogeneous Poisson process
(HPP) to the data. This is a model that assume constant intensity over
the whole space so our linear predictor is then:

$$
\eta(s) = \log\lambda(s) = \beta_0 , \ \mathbf{s}\in\Omega
$$

so the likelihood can be written as:

$$
\begin{aligned}
p(\mathbf{y} | \lambda)  & \propto \exp \left( -\int_\Omega \lambda(\mathbf{s}) \mathrm{d}\mathbf{s} \right) \prod_{i=1}^n \lambda(\mathbf{s}_i) \\
& = \exp\left( -\int_{\Omega}\exp(\beta_0)ds\right)\prod_{i=1}^n \lambda(\mathbf{s}_i) \\
\end{aligned}
$$

where $|\Omega|$ is the area of the domain of interest.

We need to approximate the integral using a numerical integration scheme
as:

$$
\approx\exp\left(-\sum_{k=1}^{N_k}w_k\lambda(s_k)\right)\prod_{i=1}^n \lambda(\mathbf{s}_i)
$$

Where $N_k$ is the number of integration points $s_1,\dots,s_{N_k}$ and
$w_1,\dots,w_{N_k}$ are the integration weights.

In this case, since the intensity is constant, the integration scheme is
really simple: it is enough to consider one random point inside the
domain with weight equal to the area of the domain.

```{r}
# define integration scheme


ips <- new_fm_int(
  st_sample(region, 1), # some random location inside the domain
  weight = st_area(region), # integration weight is the area of the domain
  name = "geometry"
)


cmp = ~ 0 + beta_0(1)

formula = geometry ~ beta_0

lik = bru_obs(data = pp,
              family = "cp",
              formula = formula,
              ips = ips)
fit1 = bru(cmp, lik)
```

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

1.  Plot the estimated posterior distribution of the intensity
2.  Compare the estimated expected number of fires on the whole domain
    with the observed ones.

`r hide("Take hint")`

Remember that in the *inlabru* framework we model the log intensity
$\eta = log(\lambda)$

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

# 1) The estimated posterior distribution of the  intensity is

post_int = inla.tmarginal(function(x) exp(x), fit1$marginals.fixed$beta_0)
post_int %>% ggplot() + geom_line(aes(x,y))

# 2) To compute the expected number of points in the area we need to multiply the
# estimated intensity by the area of the domain.
# In the same plot we also show the number of observed fires as a vertical line.

post_int = inla.tmarginal(function(x) st_area(region)* exp(x), fit1$marginals.fixed$beta_0)
post_int %>% ggplot() + geom_line(aes(x,y)) +
  geom_vline(xintercept = dim(pp)[1])

```
:::

## Fit an Inhomogeneous Poisson Process {#NHPP}

The model above has the clear disadvantages that assumes a constant
intensity and from @fig-points we clearly see that this is not the case.

The library `spatstat` contains also some covariates that can help
explain the fires distribution. Figure @fit-altitude shows the location
of fires together with the (scaled) altitude.

```{r}
#|label: fig-altitude
#|fig-cap: "Distribution of the observed forest fires and scaled altitude"
#| 
elev_raster = rast(clmfires.extra[[2]]$elevation)
elev_raster = scale(elev_raster)
ggplot() + 
  geom_spatraster(data = elev_raster) + 
  geom_sf(data = pp) +
  scale_fill_scico()

```

We are now going to use the altitude as a covariate to explain the
variability of the intensity $\lambda(s)$ over the domain of interest.

Our model is $$
\log\lambda(s) = \beta_0 + \beta_1x(s)
$$ where $x(s)$ is the altitude at location $s$.

The likelihood becomes:

$$
\begin{aligned}
p(\mathbf{y} | \lambda)  & \propto \exp \left( -\int_\Omega \lambda(\mathbf{s}) \mathrm{d}\mathbf{s} \right) \prod_{i=1}^n \lambda(\mathbf{s}_i) \\
& = \exp \left( -\int_\Omega \exp(\beta_0 + \beta_1x(s)) \mathrm{d}\mathbf{s} \right) \prod_{i=1}^n \lambda(\mathbf{s}_i) \\
\end{aligned}
$$

Now we need to choose an integration scheme to solve the integral.

In this case we will take a simple grid based approach where each
quadrature location has an equal weight. Our grid consists of
$N_k = 1000$ points and the weights are all equal to $|\Omega|/N_k$.

```{r}
#|label: fig-int2
#|fig-cap: "Integration scheme."

n.int = 1000
ips <- st_sample(region, size = n.int, type = "regular") # May not be exactly n.int points
ips <- new_fm_int(
  ips,
  weight = st_area(region) / length(ips),
  name = "geometry"
)

ggplot() + geom_sf(data = ips, aes(color = weight)) + geom_sf(data= region, alpha = 0)

```

**OBS**: The implicit assumption here is that the intensity is constant
inside each grid box, *and so is the covariate*!!

We can now fit the model:

```{r}
cmp = ~ Intercept(1) + elev(elev_raster, model = "linear")
formula = geometry ~ Intercept + elev
lik = bru_obs(data = pp,
              family = "cp",
              formula = formula,
              ips = ips)
fit2 = bru(cmp, lik)
```

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

What is the effect of the altitude on the (log) intensity of the
process?

`r hide("Take hint")`

You can look at the summary for the fixed effects

`r unhide()`

```{r}
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

fit2$summary.fixed
```
:::

::: callout-warning
⚠️ **WARNING!!**⚠️ When fitting a Point process, the integration scheme
has to be fine enough to capture the spatial variability of the
covariate!!
:::

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Rerun the model with the altitude as covariate, but this time change the
integration scheme as follows:

```{r}
n.int2 = 50
ips2 <- st_sample(region, size = n.int2, type = "regular") # May not be exactly n.int points
ips2 <- new_fm_int(
  ips2,
  weight = st_area(region) / length(ips2),
  name = "geometry"
)

```

What happens to the effect of the covariate?

`r hide("Take hint")` Re-run the model changing the integration scheme
in the *ips* input of the *bru_obs()* function.

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
#| eval: true

lik_bis = bru_obs(data = pp,
              family = "cp",
              formula = formula,
              ips = ips2)

fit2bis = bru(cmp, lik_bis)

# you can the check the differences between the two models
rbind(fit2$summary.fixed,
fit2bis$summary.fixed)

```
:::

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Now we want to predict the log-intensity over the whole domain. Use the
grid from the elevation raster to predict the intensity over the domain.

```{r}
est_grid = st_as_sf(data.frame(crds(elev_raster)), coords = c("x","y"))
est_grid  = st_intersection(est_grid, region)
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

preds2 = predict(fit2, est_grid, ~ data.frame(log_scale = Intercept + elev,

                                              lin_scale = exp(Intercept + elev)))
# then visualize it like
preds2$log_scale %>% 
  ggplot() +
  geom_sf(aes(color = mean)) +
  scale_color_scico()

```
:::

Finally, we want to use the fitted model to estimate the total number of
fires over the whole region. To do this we first have to fine the
expected number of fires as:

$$
E(N_{\Omega}) = \int_{\Omega}\exp(\lambda(s))\ ds
$$

Then simulate possible realizations of $N_{\Omega}$ to include also the
likelihood variability in our estimate:

```{r}
N_fires = generate(fit2, ips,
                      formula = ~ {
                        lambda = sum(weight * exp(elev + Intercept))
                        rpois(1, lambda)},
                    n.samples = 2000)

ggplot(data = data.frame(N = as.vector(N_fires))) +
  geom_histogram(aes(x = N),
                 colour = "blue",
                 alpha = 0.5,
                 bins = 20) +
  geom_vline(xintercept = nrow(pp),
             colour = "red") +
  theme_minimal() +
  xlab(expression(Lambda))

```

## Fit a Log-Gaussian Cox Process {#LGCP}

Finally we want to fit a LGCP with log intensity:

$$
\log(s) = \beta_0 + \beta_1x + u(s)
$$

where $\beta_0$ is the intercept, $\beta_1$ the effect of (standarized)
altitude $x(s)$ as before and $u(s)$ is a Gaussian Random field defined
through the SPDE approach.

### Define the mesh

The first step, as any time we use the SPDE approach is to defie the
mesh and the priors for the marginal variance and range:

```{r}
mesh = fm_mesh_2d(boundary = region,
                  max.edge = c(5, 10),
                  cutoff = 4, crs = NA)

ggplot() + gg(mesh) + geom_sf(data = pp)

spde_model =  inla.spde2.pcmatern(mesh,
                                  prior.sigma = c(1, 0.5),
                                  prior.range = c(100, 0.5))

```

We can then define the integration weight. Here we use the same points
to `define the SPDE approximation` and to `approximate the integral` in
the likelihood. We will see later that this does not have to be like
this, BUT integration weight and SPDE weights have to be consistent with
each other!

```{r}
ips = fm_int(mesh, samplers = region)

ggplot() + geom_sf(data = ips, aes(color = weight)) +
  gg(mesh) +
   scale_color_scico()
```

### Run the model

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Set up the components and the formula for the model above by completing
the code below and run the model.

```{r}
#| echo: true
#| eval: false


cmp = ~ ...

formula = geometry ~ ...

lik = bru_obs("cp",
              formula = formula,
              data = pp,
              ips = ...)

fit3 = bru(cmp, lik)

```

`r hide("Take hint")`

The model has three components: intercept, linear effect of altitude and
the spatial GRF `r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

cmp = ~ Intercept(1) + space(geometry, model = spde_model) + elev(elev_raster, model = "linear")

formula = geometry ~ Intercept + space + elev

lik = bru_obs("cp",
              formula = formula,
              data = pp,
              ips = ips)

fit3 = bru(cmp, lik)
```
:::

**Note** when running the model above you will get a warning:

```{r}
#| echo: false
#| eval: true
#| warning: true

fit3 = bru(cmp, lik)

```

It means that the *bru()* function cannot find the covariate values for
some of the mesh nodes. This is a common situation. As the warning says,
the *bru()* function automatically imputes the value of the covarite
using the nearest nodes. This increases the running time of the *bru()*
function, so one solution is to impute the values of the covariate over
the whole mesh 'before' running the *bru()* function.

Here, we notice that there is a single point for which elevation values
are missing (see @fig-points2 the red point that lies outside the raster
extension ).

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 4.5
#| fig-height: 4.5
#| label: fig-points2
#| fig-cap: "Integration scheme for numerical approximation of the stochastic integral in La Mancha Region"
#| purl: false
ggplot()+tidyterra::geom_spatraster(data=elev_raster)+
  geom_sf(data=ips,alpha=0.25,col=1)+
   geom_sf(data=ips[3525,],col=2)
```

To solve this, we can increase the raster extension so it covers all
both data-points and quadrature locations as well. Then, we can use the
`bru_fill_missing()` function to input the missing values with the
nearest-available-value. We can achieve this using the following code:

```{r}
#| fig-width: 4.5
#| fig-height: 4.5
#| fig-align: center
# Extend raster ext by 30 % of the original raster so it covers the whole mesh
re <- extend(elev_raster, ext(elev_raster)*1.3)
# Convert to an sf spatial object
re_df <- re %>% stars::st_as_stars() %>%  st_as_sf(na.rm=F)
# fill in missing values using the original raster 
re_df$lyr.1 <- bru_fill_missing(elev_raster,re_df,re_df$lyr.1)
# rasterize
elev_rast_p <- stars::st_rasterize(re_df) %>% rast()
ggplot() + geom_spatraster(data = elev_rast_p) 

```

::: callout-note
The `bru_fill_missing()` function was added mainly to handle very local
infilling on domain boundaries. For properly missing data, one should
consider doing a proper model of the spatial field instead.
:::

## Results

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Plot the estimated mean and standard deviation of the spatial GF and the
log-intensity over the domain of interest `r hide("Take hint")`

Use the *fm_pixels()* and *predict()* functions.

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
#| message: false
#| warning: false


pxl = fm_pixels(mesh, mask= region, dims = c(200,200))
preds = predict(fit3, pxl, ~data.frame(spde = space,
                                       log_int = Intercept + space + elev))

#and plot as 
library(scico)
library(patchwork)

ggplot(data= preds$spde) + 
  geom_sf(aes(color = mean)) + 
  scale_color_scico() +
 ggtitle("spde mean") +
ggplot(data=preds$spde ) +
  geom_sf(aes(color = sd)) +
  scale_color_scico() +
 ggtitle("spde sd") +

ggplot(data=preds$log_int) + 
  geom_sf(aes(color = mean)) + 
  scale_color_scico() +
 ggtitle("log-int mean")
ggplot(data=preds$log_int) + 
  geom_sf(aes(color = sd)) +
  scale_color_scico() +
 ggtitle("log-int sd") +
  plot_layout(ncol=2)

```
:::

Instead of just looking at the posterior mean and standard deviation, it
can be usefull to look at `simulated fields` from the posterior
distribution. This is because the mean field is, by definition, smoother
than any realization of the field. So looking at simulation can give us
a better idea of how the field might look like. We can do this using the
*generate()* function:

```{r}
sim_fields = generate(fit3, pxl, ~data.frame(spde = space,
                                       log_int = Intercept + space + elev),
                     n.samples = 4)

cbind(pxl,sapply(sim_fields, function(x) x$spde)) %>%
  pivot_longer(-geometry) %>%
  ggplot() + geom_sf(aes(color = value)) + 
  facet_wrap(.~name) + scale_color_scico() +
  ggtitle("simulated spatial fields")


cbind(pxl,sapply(sim_fields, function(x) x$log_int)) %>%
  pivot_longer(-geometry) %>%
  ggplot() + geom_sf(aes(color = value)) + 
  facet_wrap(.~name) + scale_color_scico() + 
  ggtitle("simulated log intensity")


```
