---
title: "Space-time modeling"
execute: 
  warning: false
  message: false
format: 
  html:
    theme:
      light: flatly
      dark: darkly
  PrettyPDF-pdf:
    keep-tex: true
    number-sections: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| message: false
#| warning: false
#| purl: false

# load webexercises library for tasks and questions (just for a preview - the practical compiler should take care of this when compiling multiple excercises)
library(webexercises)

```

------------------------------------------------------------------------

Libraries to load:

```{r}
#| warning: false
#| message: false


library(dplyr)
library(INLA)
library(inlabru) 
library(sf)
library(terra)


# load some libraries to generate nice map plots
library(scico)
library(ggplot2)
library(patchwork)
library(mapview)
library(tidyterra)

```

### The data

In this practical, we will revisit the data on the Pacific Cod (*Gadus macrocephalus*) from a trawl survey in Queen Charlotte Sound. The `pcod` dataset is available from the `sdmTMB` package and contains the presence/absence records of the Pacific Cod during each surveys along with the biomass density of Pacific cod in the area swept (kg/Km$^2$). The `qcs_grid` data contain the depth values stored as $2\times 2$ km grid for Queen Charlotte Sound.

The dataset contains presence/absence data from 2003 to 2017. To make computations faster we only consider the first 3 years.

```{r}
#| message: false
#| warning: false
library(sdmTMB)

pcod_df = sdmTMB::pcod  %>% filter(year<=2005)
qcs_grid = sdmTMB::qcs_grid

```

Then, we create a `sf` object and assign the right coordinate reference to it:

```{r}
#| message: false
#| warning: false
#| 
pcod_sf =   st_as_sf(pcod_df, coords = c("lon","lat"), crs = 4326)
pcod_sf = st_transform(pcod_sf,
          crs = "+proj=utm +zone=9 +datum=WGS84 +no_defs +type=crs +units=km" )
```

We convert the covariates into a raster and assign the same coordinate reference:

```{r}

depth_r <- rast(qcs_grid, type = "xyz")
crs(depth_r) <- crs(pcod_sf)

```

```{r}
#| echo: false
#| purl: false
#| fig-align: center
#| fig-width: 8
#| fig-height: 8

theme_map = theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())

ggplot()+ 
  geom_spatraster(data=depth_r$depth)+
  geom_sf(data=pcod_sf,aes(color=factor(present))) +
    scale_color_manual(name="Occupancy status for the Pacific Cod",
                     values = c("black","orange"),
                     labels= c("Absence","Presence"))+
  scale_fill_scico(name = "Depth",
                   palette = "nuuk",
                   na.value = "transparent" ) + xlab("") + ylab("") + facet_wrap(~year) + theme_map

```

### Spatio-temporal modeling

#### Model fitting

Now lets compare two different space-time models using LGOCV and some information criteria metrics. The general model structure is given by:

$$
\begin{aligned}
y(s,t)|\eta(s,t)&\sim\text{Binom}(1, p(s,t))\\
\eta(s,t) &= \text{logit}(p(s,t)) \\
\end{aligned}
$$ We also want to compare the models using WAIC, DIC and marginal likelihood:

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Set the `bru_options` so that the quantities of interest are computed

```{r}
#| purl: false
#| webex.hide: "Click here to see the solution"
#| warning: false
#| message: false
#| eval: true
#| 
bru_options_set(control.compute = list(waic = TRUE,dic= TRUE,mlik = TRUE))
```
:::

#### Model 1

1.  **Model 1 - time iid effect** We consider a separable space-time model with a linear predictors given by:

$$
\eta(s,t) = \beta_0 + f_1(\text{depth}(s)) + f_2(t) + \omega(s)
$$

-   $f_1(\text{depth}(s))$ is a smooth covariate effect of depth (modeled using a RW2 model)

-   $f_2(t)$ is an IID effect of time

-   $\omega(s)$ is Matérn random field.

The first step is to define the mesh and the spde model

**Construct the mesh and the SPDE model**

```{r}
mesh = fm_mesh_2d(loc = pcod_sf,    
                  cutoff = 1,
                  max.edge = c(10,20),     
                  offset = c(5,50),
                  crs = st_crs(pcod_df))   

spde_model =  inla.spde2.pcmatern(mesh,
                                  prior.sigma = c(1, 0.5),
                                  prior.range = c(100, 0.5))
```

```{r}
#| echo: false
#| purl: false
#| fig-align: center
#| fig-width: 5
#| fig-height: 5
ggplot() + gg(mesh) + geom_sf(data= pcod_sf, aes(color = factor(present)), size = 0.1) + xlab("") + ylab("")

```

**create time index and the grouped variable**

To use the RW2 model the covariate has to be grouped:

```{r}
depth_r$depth_group = inla.group(values(depth_r$depth_scaled))

```

we also define a time index from 1 to in the data frame

```{r}
pcod_sf = pcod_sf %>%
     mutate(time_idx = match(year, c(2003, 2004, 2005)),
         id = 1:nrow(.)) # Observation id for CV
```

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Implement the model in `inlabru`

1.  Define the components
2.  Define the formula
3.  Define the likelihood model using the `bru_obs()` function
4.  Run the model

```{r}
#| purl: false
#| webex.hide: "Click here to see the solution"
#| warning: false
#| message: false
#| eval: true

# Model components
cmp_spat = ~ Intercept(1) + 
  covariate(depth_r$depth_group, model = "rw2", scale.model = TRUE, 
           values = sort(unique(values(depth_r$depth_group)))
            )+
  trend(time_idx, model = "iid")+
  space(geometry, model = spde_model)

# Linear predictor
formula_spat = present ~ Intercept  + trend + space + covariate

# Observational model
lik_spat = bru_obs(formula = formula_spat, 
              data = pcod_sf, 
              family = "binomial")

# Fit Model 
fit_spat = bru(cmp_spat,lik_spat)


```
:::

::: callout-note
Note that there are some survey locations in certain years that fall outside the depth raster region. `inlabru` will input these missing covariate values using the nearest available value. This can be computationally expensive, but you can avoid it by supplying a raster layer that encompasses all of your data points (e.g., by pre-imputing these missing values with your preferred method of choice).

One way of doing this in the `inlabru` framework is to use the `bru_fill_missing()` function:

```{r}
# Select the raster of interest
depth_orig = depth_r$depth_group
re <- extend(depth_orig, ext(depth_orig)*1.05)
# Convert to an sf spatial object
re_df <- re %>% stars::st_as_stars() %>%  st_as_sf(na.rm=F)
# fill in missing values using the original raster 
re_df$depth_group =  bru_fill_missing(depth_orig,re_df,re_df$depth_group)
# rasterize
depth_filled <- stars::st_rasterize(re_df) %>% rast()
plot(depth_filled)
```
:::

We have now fit the model and want to check the results.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Is the effect of depth significant? Does it seem important to have a non-linear model?

Inspect the estimated time effect and

Use the `predict` function to inspect the estimated spatial effect.

Use the `depth_r` raster to define the points in space where to predict

```{r}
pxl = st_as_sf(data.frame(crds(depth_r)), coords = c("x","y") ,
               crs  = st_crs(pcod_sf))
```

```{r}
#| purl: false
#| webex.hide: "Click here to see the solution"
#| warning: false
#| message: false
#| eval: true

# covariate effect

p_cov = fit_spat$summary.random$covariate %>%
  ggplot() + geom_ribbon(aes(ID, ymin = `0.025quant`, ymax = `0.975quant` )) +
  geom_line(aes(ID,mean)) + ggtitle("Covariate effect")

# time effect
p_time = fit_spat$summary.random$trend %>%
  ggplot() + geom_errorbar(aes(ID, ymin = `0.025quant`, ymax = `0.975quant` )) +
  geom_point(aes(ID,mean)) + ggtitle("Time effect")

#space effect
pred_space = predict(fit_spat, pxl, ~ space)

p_space_mean = ggplot() + gg(pred_space, aes(color = mean))
p_space_sd = ggplot() + gg(pred_space, aes(color = sd))

```
:::

#### Model 2

1.  **Model 2 - spatiotemporal field** We consider a separable space time model with a linear predictor given by:

$$
\eta(s,t) = \beta_0 + f_1(\text{depth}(s)) + \omega(s,t)
$$

-   $f_1(\text{depth}(s))$ is a smooth covariate effect of depth (RW2)
-   $\omega(s,t)$ is a space-time Matérn spatial field with AR1 time component $$
    \omega(s,t) = \phi\ \omega(s,t-1) + \epsilon(s),\qquad \epsilon(s)\sim\text{GF}(\sigma_{\epsilon},\rho_{\epsilon})
    $$

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Implement this second model in `inlabru`

1.  Define the components For the AR1 model use this following PC prior for the correlation parameter $\phi$

```{r}
# PC prior for AR(1) correlation parameter
h.spec <- list(rho = list(prior = 'pc.cor0', param = c(0.5, 0.1)))
```

2.  Define the formula
3.  Define the likelihood model using the `bru_obs()` function
4.  Run the model (This model can take a couple of minutes to run)

```{r}
#| purl: false
#| webex.hide: "Click here to see the solution"
#| warning: false
#| message: false
#| eval: true

# Model components
cmp_spat_ar1 = ~ Intercept(1) + 
  covariate(depth_filled$depth_group, model = "rw2", scale.model = TRUE)+
  space_time(geometry,
        group = time_idx ,
        model = spde_model,
        control.group = list(model = 'ar1',hyper = h.spec))

# Linear predictor
formula_spat_ar1 = present ~ .

# Observational model
lik_spat_ar1 = bru_obs(formula = formula_spat_ar1, 
              data = pcod_sf, 
              family = "binomial")

# Fit Model 
fit_spat_ar1 = bru(cmp_spat_ar1,lik_spat_ar1)

```
:::

Now we want to check the results

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

What is the estimated parameter $phi$ in the auto-regressive part of the model?

Check the effect of the covariate.

Use the `predict` function to inspect the estimated probability of presence. Use the same prediction points as before, but here you also need to use the \`

```{r}
#| purl: false
#| webex.hide: "Click here to see the solution"
#| warning: false
#| message: false
#| eval: true

# autoregressive effect

#fit_spat_ar1$summary.hyperpar

#covariate
p_cov_ar1 = fit_spat_ar1$summary.random$covariate %>%
  ggplot() + geom_ribbon(aes(ID, ymin = `0.025quant`, ymax = `0.975quant` )) +
  geom_line(aes(ID,mean)) + ggtitle("Covariate effect")


#space-time effect

inv_logit = function(x){ exp(x) / (1 + exp(x))}

pxl_all = fm_cprod(pxl, data.frame(time_idx = 1:3))
pred_space_ar1 = predict(fit_spat_ar1, pxl_all, ~data.frame(logit_prob = Intercept  + 
                                                               covariate  + 
                                                               space_time ,
                                                            prob = inv_logit(Intercept  + 
                                                               covariate  + 
                                                               space_time)))


p_ar1 = pred_space_ar1$prob %>% ggplot() + geom_sf(aes(color = mean)) + 
  facet_wrap(.~time_idx) + scale_color_scico(direction = -1) +
  theme_map


```
:::

### Model Comparison

Now we want to use the WAIC, DIC and MLIK to compare the models

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Compare the scores, what is your conclusion?

```{r}
#| purl: false
#| webex.hide: "Click here to see the solution"
#| warning: false
#| message: false
#| eval: true

out= data.frame(Model = c("Model 1", "Model 2"),
  DIC = c(fit_spat$dic$dic, fit_spat_ar1$dic$dic),
  WAIC = c(fit_spat$waic$waic, fit_spat_ar1$waic$waic),
  MLIK = c(fit_spat$mlik[1], fit_spat_ar1$mlik[1]))

```
:::
