---
title: "Practical - Dealing with excess zeros"
execute: 
  warning: false
  message: false
format: 
  html:
    theme:
      light: flatly
      dark: darkly
  PrettyPDF-pdf:
    keep-tex: false
    number-sections: true
embed-resources: true
number-sections: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
# bibliography: zip_zap_models.bib
---

```{r}
#| echo: false
#| include: false
#| purl: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5
)

# load webexercises library for tasks and questions (just for a preview - the practical compiler should take care of this when compiling multiple excercises)
library(webexercises)
```

# ZIP/ZAP and Hurdle Models

```{r setup}
#| eval: true
#| echo: true
#| include: true
#| message: false
#| warning: false
#| label: setup

library(dplyr)
library(ggplot2)
library(inlabru)
library(INLA)
library(terra)
library(sf)
library(scico)
library(magrittr)
library(patchwork)
library(tidyterra)


# We want to obtain CPO data from the estimations
bru_options_set(control.compute = list(dic = TRUE,
                                       waic = TRUE,
                                       mlik = TRUE,
                                       cpo = TRUE))
```

In this practical we are going to work with data with excess zeros. We
will

-   [Create count data from a `gorillas` dataset](#sec-prep)

-   Fit a [zero inflated model](#sec-zip)

-   Fit a [hurdle model](#sec-zap)

-   Fit a [hurdle model using two likelihoods](#sec-two-lik)

-   Fit a [hurdle model using two likelihoods and a shared
    component](#sec-two-lik-share)

## Data Preparation {#sec-prep}

The following example use the `gorillas` dataset available in the
`inlabru` library.

The data give the locations of Gorilla's nests in an area:

```{r }
#| fig-cap: "Location of gorilla nests"
#| out-width: "80%"
#| fig-align: 'center'
#| label: nests_loc
gorillas_sf <- inlabru::gorillas_sf
nests <- gorillas_sf$nests
boundary <- gorillas_sf$boundary

ggplot() + geom_sf(data = nests) +
  geom_sf(data = boundary, alpha = 0)

```

The dataset also contains covariates in the form or raster data. We
consider two of them here:

```{r}
gcov = gorillas_sf_gcov()
elev_cov <- gcov$elevation
dist_cov <-  gcov$waterdist
```

```{r}
#| fig-cap: "Covariates"
#| out-width: "80%"
#| fig-align: 'center'
#| echo: false
#| 
theme_map = theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())

p1 = ggplot() + geom_spatraster(data = elev_cov) + ggtitle("Elevation") +
  scale_fill_scico(direction = -1) + geom_sf(data = boundary, alpha = 0) + theme_map+ theme(legend.position = "none")
p2 = ggplot() + geom_spatraster(data = dist_cov) + ggtitle("Distance to water") +
  scale_fill_scico(direction = -1) + geom_sf(data = boundary, alpha = 0) +
  theme_map + theme(legend.position = "none")
  
p1 + p2
```

**Note:** the covariates have been expanded to cover all the nodes in
the mesh.

------------------------------------------------------------------------

To obtain the count data, we rasterize the species counts to match the
spatial resolution of the covariates available. Then we aggregate the
pixels to a rougher resolution (5x5 pixels in the original covariate
raster dimensions). Finally, we mask regions outside the study area.

In addition we compute the area of each grid cell.

```{r}
#| fig-cap: "Counts of gorilla nests"
#| out-width: "70%"
#| fig-align: 'center'
#| 
# Rasterize data
counts_rstr <-
  terra::rasterize(vect(nests), gcov, fun = sum, background = 0) %>%
  terra::aggregate(fact = 5, fun = sum) %>%
  mask(vect(sf::st_geometry(boundary)))
plot(counts_rstr)
# compute cell area
counts_rstr <- counts_rstr %>%
  cellSize(unit = "km") %>%
  c(counts_rstr)
```

To create our dataset of counts, we extract also the coordinate of
center point of each raster pixel. In addition we create a column with
presences and one with the pixel area

```{r}
#| echo: true
counts_df <- crds(counts_rstr, df = TRUE, na.rm = TRUE) %>%
  bind_cols(values(counts_rstr, mat = TRUE, na.rm = TRUE)) %>%
  rename(count = sum) %>%
  mutate(present = (count > 0) * 1L) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(nests))
```

We then aggregate the covariates to the same resolution as the nest
counts and scale them.

```{r}
elev_cov1 <- elev_cov %>% 
  terra::aggregate(fact = 5, fun = mean) %>% scale()
dist_cov1 <- dist_cov %>% 
  terra::aggregate(fact = 5, fun = mean) %>% scale()
```

```{r}
#| label: fig-covariate-raster
#| fig-cap: "Covariates "
#| out-width: "80%"
#| fig-align: 'center'
#| echo: false

p1 = ggplot() + geom_spatraster(data = elev_cov1) + ggtitle("Elevation") +
  scale_fill_scico(direction = -1) + geom_sf(data = boundary, alpha = 0) + theme_map
p2 = ggplot() + geom_spatraster(data = dist_cov1) + ggtitle("Distance to water") +
  scale_fill_scico(direction = -1) + geom_sf(data = boundary, alpha = 0) +
  theme_map
  
p1 + p2

```

### Mesh building

We now define the mesh and the spde object.

```{r}

mesh <- fm_mesh_2d(
  loc = st_as_sfc(counts_df),
  max.edge = c(0.5, 1),
  crs = st_crs(counts_df)
)

matern <- inla.spde2.pcmatern(mesh,
  prior.sigma = c(1, 0.01),
  prior.range = c(5, 0.01)
)
```

```{r}
#| fig-cap: "Mesh over the count locations"
#| fig-align: 'center'
#| out-width: "80%"
#| echo: false
ggplot() +
  geom_fm(data = mesh) +
  geom_sf(
    data = counts_df[counts_df$count > 0, ],
    aes(color = count),
    size = 1,
    pch = 4
  ) +
  theme_minimal() + theme_map
```

In our dataset, the number of zeros is quite substantial, and our model
may struggle to account for them adequately. To address this, we should
select a model capable of handling an "inflated" number of zeros,
exceeding what a standard Poisson model would imply. For this purpose,
we opt for a "zero-inflated Poisson model," commonly abbreviated as ZIP.

## Zero-Inflated model (Type1) {#sec-zip}

We fit now a Zero-Inflated model to our data.

The [Type 1 Zero-inflated Poisson
model](https://inla.r-inla-download.org/r-inla.org/doc/likelihood/zeroinflated.pdf)
is defined as follows:

$$
\text{Prob}(y\vert\dots)=\pi\times 1_{y=0}+(1-\pi)\times \text{Poisson}(y)
$$

Here, $\pi=\text{logit}^{-1}(\theta)$

The expected value and variance for the counts are calculated as:

$$
\begin{gathered}
E(count)=(1-\pi)\lambda \\
Var(count)= (1-\pi)(\lambda+\pi \lambda^2)
\end{gathered}
$$ {#eq-mean-zip}

This model has two parameters:

-   The probability of excess zero $\pi$ - This is a *hyperparameter*
    and therefore it is constant
-   The mean of the Poisson distribution $\lambda$. This is linked to
    the linear predictor as: $$
    \eta = E\log(\lambda) = \log(E) + \beta_0 + \beta_1\text{Elevation} + \beta_2\text{Distance } + u
    $$ where $\log(E)$ is an offset (the area of the pixel) that
    accounts for the size of the cell.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Fit a zero-inflated model to the data (`zeroinflatedpoisson1`) by
completing the following code:

```{r}
#| echo: true
#| eval: false

cmp = ~ Intercept(1) + elevation(...) + distance(...) + space(...)

lik = bru_obs(...,
    E = area)

fit_zip <- bru(cmp, lik)
```

`r hide("Take hint")` The `E = area` is an offset that adjusts for the
size of each cell.

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false


cmp = ~ Intercept(1) + elevation(elev_cov1, model = "linear") + distance(dist_cov1, model = "linear") + space(geometry, model = matern)



lik = bru_obs(formula = count ~ .,
    family = "zeroinflatedpoisson1", 
    data = counts_df,
    E = area)

fit_zip <- bru(cmp, lik)
```
:::

Once the model is fitted we can look at the results

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Check what the estimated excess zero probaility is.

Use the `predict()` function to look at the estimated $\lambda(s)$ and
mean count in @eq-mean-zip

`r hide("Take hint")`

To get the right name for the hyperparameters to use in the `predict()`
function, you can use the function `bru_names()`.

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

# to check the estimated excess zero probability:
# fit_zip$summary.hyperpar

pred_zip <- predict(
  fit_zip, 
  counts_df,
  ~ {
    pi <- zero_probability_parameter_for_zero_inflated_poisson_1
    lambda <- area * exp( distance + elevation + space + Intercept)
    expect <- (1-pi) * lambda
    variance <- (1-pi) * (lambda + pi * lambda^2)
    list(
      lambda = lambda,
      expect = expect,
      variance = variance
    )
  },
  n.samples = 2500
)
```
:::

```{r}
#| fig-cap: "Estimated $\\lambda$ (left) and expected counts (right) with zero inflated model"
#| fig-align: 'center'
#| out-width: "80%"
#| echo: false
p1 = ggplot() + gg(pred_zip$lambda, aes(fill = mean), geom="tile") + theme_map + scale_fill_scico(direction = -1) + 
  ggtitle(expression("Posterior mean of " ~lambda)) + theme(legend.position = "bottom")
p2 = ggplot() + gg(pred_zip$expect, aes(fill = mean), geom="tile") + theme_map + scale_fill_scico(direction = -1)+ 
  ggtitle(expression("Posterior mean of Expected counts"))+ theme(legend.position = "bottom")
p1 + p2

```

## Hurdle model (Type0) {#sec-zap}

We now fit a hurdle model to the same data.

In the `zeroinflatedpoisson0` model is defined by the [following
observation probability
model](https://inla.r-inla-download.org/r-inla.org/doc/likelihood/zeroinflated.pdf)

$$
\text{Prob}(y\vert\dots)=\pi\times 1_{y=0}+(1-\pi)\times \text{Poisson}(y\vert y>0)
$$

where $\pi$ is the probability of zero.

The expectation and variance of the counts are as follows:

$$
\begin{aligned}
E(\text{count})&=\frac{1}{1-\exp(-\lambda)}\pi\lambda \\
Var(\text{count})&=  E(\text{count}) \left(1-\exp(-\lambda) E(\text{count})\right)
\end{aligned}
$$ {#eq-mean-zap}

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Fit a hurdle model to the data using the `zeroinflatedpoisson0`
likelihood `r hide("Take hint")`

You do not need to redefine the components as the linear predictor is
not changing.

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

lik = bru_obs(formula = count ~ .,
    family = "zeroinflatedpoisson0", 
    data = counts_df,
    E = area)

fit_zap <- bru(cmp, lik)

```
:::

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

As before, check what the estimated probability of zero is and use
`predict()` to obtain a map of the estimated mean counts in @eq-mean-zap
over the domain.

`r hide("Take hint")`

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false


pred_zap <- predict( fit_zap, counts_df,
  ~ {
    pi <- zero_probability_parameter_for_zero_inflated_poisson_0
    lambda <- area * exp( distance + elevation + space + Intercept)
    expect <- ((1-exp(-lambda))^(-1) * pi * lambda)
    list(
      lambda = lambda,
      expect = expect
    )
  },
  n.samples = 2500
)



```
:::

```{r}
#| fig-cap: "Estimated $\\lambda$ (left) and expected counts (right) with hurdle model"
#| fig-align: 'center'
#| out-width: "80%"
#| echo: false
p1 = ggplot() + gg(pred_zap$lambda, aes(fill = mean), geom="tile") + theme_map + scale_fill_scico(direction = -1) + 
  ggtitle(expression("Posterior mean of " ~lambda)) + theme(legend.position = "bottom")
p2 = ggplot() + gg(pred_zap$expect, aes(fill = mean), geom="tile") + theme_map + scale_fill_scico(direction = -1)+ 
  ggtitle(expression("Posterior mean of Expected counts"))+ theme(legend.position = "bottom")
p1 + p2

```

## Hurdle model using two likelihoods {#sec-two-lik}

Here the model is the same as in @sec-zap, but this time we also want to
model $\pi$ using covariates and random effects. Therefore we define a
second linear predictor $$
\eta^2 =\beta_0^2 + \beta_1^2\text{Elevation} +  \beta_2^2\text{Distance} + u^2 
$$ **Note** here we have defined the two linear predictor to use the
same covariates, but this is not necessary, they can be totally
independent.

To fit this model we have to define two likelihoods: - One will account
for the presence-absence process and has a Binomial model - One will
account for the counts and has a truncated Poisson model

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Complete the following code to fit a hurdle model based on two
likelihoods:

```{r}
#| echo: true
#| eval: false

# define components
cmp <- ~
  Intercept_count(1) +
    elev_count(elev_cov1, model = "linear") +
    dist_count(dist_cov1, model = "linear") +
    space_count(geometry, model = matern) +
    Intercept_presence(1) +
    elev_presence(elev_cov1, model = "linear") +
    dist_presence(dist_cov1, model = "linear") +
    space_presence(geometry, model = matern) 

# positive count model
pos_count_obs <- bru_obs(formula = ...,
      family = ...,
      data = counts_df[counts_df$present > 0, ],
      E = area)
  
# presence model
presence_obs <- bru_obs(formula ...,
  family = ...,
  data = counts_df,
)

# fit the model
fit_zap2 <- bru(...)
```

`r hide("Take hint")`

Add hint details here...

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

cmp <- ~
  Intercept_count(1) +
    elev_count(elev_cov1, model = "linear") +
    dist_count(dist_cov1, model = "linear") +
    space_count(geometry, model = matern) +
    Intercept_presence(1) +
    elev_presence(elev_cov1, model = "linear") +
    dist_presence(dist_cov1, model = "linear") +
    space_presence(geometry, model = matern) 


pos_count_obs <- bru_obs(formula = count ~ Intercept_count + elev_count + 
                                   dist_count + space_count,
      family = "nzpoisson",
      data = counts_df[counts_df$present > 0, ],
      E = area)
  

presence_obs <- bru_obs(formula = present ~ Intercept_presence + elev_presence + dist_presence +
                          space_presence,
  family = "binomial",
  data = counts_df,
)

fit_zap2 <- bru(
  cmp,
  presence_obs,
  pos_count_obs
)

```
:::

## Hurdle model using two likelihoods and a shared component {#sec-two-lik-share}

Note that in the model above, there is no direct link between the
parameters of the two observation parts, and we could estimate them
separately. However, the two likelihoods could share some of the
components; for example the `space_count` component could be used for
both predictors. This would be possible using the `copy` argument.

We would then need to define one component as
`space(geometry, model = matern)` and then a copy of it as
`space_copy(geometry, copy = "space", fixed = FALSE)`.

The results from the model in @sec-sec-two-lik show that the estimated
covariance parameters for the two fields are very different, so it is
probably not sensible to share the same component between the two parts.
We do it anyway to show an example:

```{r fig-fit-zap}
#| eval: true
#| echo: true

cmp <- ~
  Intercept_count(1) +
    elev_count(elev_cov1, model = "linear") +
    dist_count(dist_cov1, model = "linear") +
    Intercept_presence(1) +
    elev_presence(elev_cov1, model = "linear") +
    dist_presence(dist_cov1, model = "linear") +
    space(geometry, model = matern) +
  space_copy(geometry, copy = "space", fixed = FALSE)


pos_count_obs <- bru_obs(formula = count ~ Intercept_count + elev_count + dist_count + space,
      family = "nzpoisson",
      data = counts_df[counts_df$present > 0, ],
      E = area)

presence_obs <- bru_obs(formula = present ~ Intercept_presence + elev_presence + dist_presence + space_copy,
  family = "binomial",
  data = counts_df)

fit_zap3 <- bru(
  cmp,
  presence_obs,
  pos_count_obs)
```

## Comparing models

We have fitted four different models. Now we want to compare them and
see how they fit the data.

### Comparing model predictions

We first want to compare the estimated surfaces of expected counts. To
do this we want to produce the estimated expected counts, similar to
what we did in @sec-zip and @sec-zap for all four models and plot them
together:

```{r}
#| fig-cap: "Estimated expected counts for all four models"
#| out-width: 80%
pred_zip <- predict(
  fit_zip, 
  counts_df,
  ~ {
    pi <- zero_probability_parameter_for_zero_inflated_poisson_1
    lambda <- area * exp( distance + elevation + space + Intercept)
    expect <- (1-pi) * lambda
    variance <- (1-pi) * (lambda + pi * lambda^2)
    list(
      expect = expect
    )
  },n.samples = 2500)

pred_zap <- predict( fit_zap, counts_df,
  ~ {
    pi <- zero_probability_parameter_for_zero_inflated_poisson_0
    lambda <- area * exp( distance + elevation + space + Intercept)
    expect <- ((1-exp(-lambda))^(-1) * pi * lambda)
    list(
      expect = expect)
  },n.samples = 2500)

inv.logit = function(x) (exp(x)/(1+exp(x)))

pred_zap2 <- predict( fit_zap2, counts_df,
  ~ {
    pi <- inv.logit(Intercept_presence + elev_presence + dist_presence + space_presence)
    lambda <- area * exp( dist_count + elev_count + space_count + Intercept_count)
    expect <- ((1-exp(-lambda))^(-1) * pi * lambda)
    list(
      expect = expect)
  },n.samples = 2500)

pred_zap3 <- predict( fit_zap3, counts_df,
  ~ {
    pi <- inv.logit(Intercept_presence + elev_presence + dist_presence + space_copy)
    lambda <- area * exp( dist_count + elev_count + space + Intercept_count)
    expect <- ((1-exp(-lambda))^(-1) * pi * lambda)
    list(
      expect = expect)
  },n.samples = 2500)




p =   data.frame(x = st_coordinates(counts_df)[,1],
             y = st_coordinates(counts_df)[,2],
    zip = pred_zip$expect$mean,
         hurdle = pred_zap$expect$mean,
         hurdle2 = pred_zap2$expect$mean,
         hurdle3 = pred_zap3$expect$mean)  %>%
  pivot_longer(-c(x,y)) %>%
  ggplot() + geom_tile(aes(x,y, fill = value)) + facet_wrap(.~name) +
    theme_map + scale_fill_scico(direction = -1)




```

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Create plots of the estimated variance of the counts.

`r hide("Take hint")`

The fomulas for the variances are in @eq-mean-zip and @eq-mean-zap.

`r unhide()`

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

pred_zip <- predict(
  fit_zip, 
  counts_df,
  ~ {
    pi <- zero_probability_parameter_for_zero_inflated_poisson_1
    lambda <- area * exp( distance + elevation + space + Intercept)
    variance <- (1-pi) * (lambda + pi * lambda^2)
    list( variance = variance)
  },n.samples = 2500)

pred_zap <- predict( fit_zap, counts_df,
  ~ {
    pi <- zero_probability_parameter_for_zero_inflated_poisson_0
    lambda <- area * exp( distance + elevation + space + Intercept)
    expect <- ((1-exp(-lambda))^(-1) * pi * lambda)
    variance = expect *(1-exp(-lambda) * expect)
    list(variance = variance)
  },
  n.samples = 2500)

inv.logit = function(x) (exp(x)/(1+exp(x)))

pred_zap2 <- predict( fit_zap2, counts_df,
  ~ {
    pi <- inv.logit(Intercept_presence + elev_presence + dist_presence + space_presence)
    lambda <- area * exp( dist_count + elev_count + space_count + Intercept_count)
    expect <- ((1-exp(-lambda))^(-1) * pi * lambda)
    variance = expect *(1-exp(-lambda) * expect)
    list(variance = variance)
  },
  n.samples = 2500)

pred_zap3 <- predict( fit_zap3, counts_df,
  ~ {
    pi <- inv.logit(Intercept_presence + elev_presence + dist_presence + space_copy)
    lambda <- area * exp( dist_count + elev_count + space + Intercept_count)
    expect <- ((1-exp(-lambda))^(-1) * pi * lambda)
    variance = expect *(1-exp(-lambda) * expect)
    list(variance = variance)
  },
  n.samples = 2500)




  data.frame(x = st_coordinates(counts_df)[,1],
             y = st_coordinates(counts_df)[,2],
    zip = pred_zip$variance$mean,
         hurdle = pred_zap$variance$mean,
         hurdle2 = pred_zap2$variance$mean,
         hurdle3 = pred_zap3$variance$mean)  %>%
  pivot_longer(-c(x,y)) %>%
  ggplot() + geom_tile(aes(x,y, fill = value)) + facet_wrap(.~name) +
    theme_map + scale_fill_scico(direction = -1)


  
```
:::

### Using scores

We can compare model using the scores that the *bru()* function computes
since we have set, at the beginning. the options to

```{r}
#| echo: true
#| eval: false

bru_options_set(control.compute = list(dic = TRUE,
                                       waic = TRUE,
                                       mlik = TRUE,
                                       cpo = TRUE))

```

Lets use these scores to compare the models.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Extract the DIC, WAIC and MLIK values for the four models and compare
them

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

data.frame( Model = c("ZIP", "HURDLE", "HURDLE_2","HURDLE_3" ),
  DIC = c(fit_zip$dic$dic, fit_zap$dic$dic, WAIC = fit_zap2$dic$dic, fit_zap3$dic$dic),
            WAIC = c(fit_zip$waic$waic, fit_zap$waic$waic, fit_zap2$waic$waic, fit_zap3$waic$waic),
            MLIK = c(fit_zip$mlik[1], fit_zap$mlik[1], fit_zap2$mlik[1], fit_zap3$mlik[1]))
            


```
:::

From the table above we can see that the model that best balances
complexity and fit is the zero inflated one (ZIP).
